---
title: "Introduction to fiaplyr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to fiaplyr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette serves as a general introduction to `fiaplyr`, a package designed to make working with Forest Inventory and Analysis (FIA) databases simpler and more intuitive. It covers the core concepts, from connecting to a database to producing population estimates.

## Introduction to FIA and `fiaplyr`

The Forest Inventory and Analysis (FIA) program is often referred to as the "nation's forest census." It collects data on a grid of plots across the United States. While the data is rich, the database structure is complex, involving many tables (Plots, Conditions, Trees, etc.) and intricate sampling designs.

`fiaplyr` provides a modern interface for this data. Instead of writing complex SQL queries to join tables and filter for specific inventories, `fiaplyr` allows you to express your analysis in R using familiar `dplyr`-style verbs, while handling the database logic behind the scenes.

## The `fiaplyr` Philosophy

Two main principles guide `fiaplyr`:

1.  **Evaluation-Centric**: FIA data is organized into "Evaluations" (identified by an EVALID). An Evaluation represents a specific population of interest (e.g., "Oregon 2010"). `fiaplyr` handles the complex filtering required to respect these boundaries automatically.
2.  **Lazy Evaluation**: `fiaplyr` is built on `dbplyr`. This means that when you write R code to manipulate data, you aren't pulling millions of rows into memory. Instead, `fiaplyr` translates your commands into SQL and executes them in the database. Data is only brought into R when you explicitly ask for it (e.g., using `collect()`).

## Setup

To use `fiaplyr`, you first need a connection to an FIA database (e.g., SQLite, DuckDB, PostgreSQL).

```{r setup, eval=FALSE}
library(fiaplyr)
library(dplyr)
library(DBI)
library(duckdb) # or RSQLite, RPostgres, etc.

# Establish a connection to your database
db_path <- "./path/to/fiadb.duckdb"
con <- dbConnect(duckdb::duckdb(), db_path)
```

## The Evaluation Handler

The core object in `fiaplyr` is the **Evaluation Handler**. You create one by specifying your database connection and an Evaluation ID (EVALID). This object acts as your interface to the data for that specific inventory.

```{r handler, eval=FALSE}
# Initialize handler for a specific EVALID (e.g., 411001 for Oregon 2010)
handler <- eval_handler(con, evalid = 411001)

# Inspect the handler to see what data it covers
print(handler)
```

The handler automatically sets up "lazy" tables for the Plot, Condition, and Tree data relevant to that evaluation.

## Basic Analysis

You can perform plot-level aggregations directly on the handler. `fiaplyr` provides a formula interface (`slot ~ variable`) to specify what you want to calculate.

For example, to calculate the Net Cubic Foot Volume (`VOLCFNET`) per plot:

```{r aggregate, eval=FALSE}
# Calculate volume per acre for each plot
# 'tree' is the slot (source table), 'VOLCFNET' is the variable
plot_vol <- handler |>
  aggregate(tree ~ VOLCFNET)

# The result is a lazy query. Use collect() to retrieve the data.
collect(plot_vol)
```

`fiaplyr` automatically handles the weighting (multiplying by `TPA_UNADJ` or trees per acre) so you get per-acre estimates by default.

## Population Estimation

While plot-level summaries are useful, the primary goal of FIA analysis is often to estimate totals for a population (e.g., "Total Volume in Oregon"). This requires a **Post-Stratified Estimator**.

`fiaplyr` simplifies this process with the `PostStratifiedEstimator` class.

```{r estimation, eval=FALSE}
# Create an estimator from your handler
estimator <- PostStratifiedEstimator(handler)

# Estimate total Net Cubic Foot Volume for the entire population
vol_est <- estimate(estimator, tree ~ VOLCFNET)

print(vol_est)
```

### Domain Estimation

You can also calculate estimates for specific groups (domains). This is done by setting domains on the handler before creating the estimator.

```{r domain_est, eval=FALSE}
# Estimate Volume by Species
vol_by_sp_est <- handler |>
  set_tree_domains(SPCD) |>        # Group by Species Code
  PostStratifiedEstimator() |>     # Create estimator with these groups
  estimate(tree ~ VOLCFNET)        # Calculate estimate

collect(vol_by_sp_est)
```

## Conclusion

This vignette covered the basics of connecting to an FIA database, selecting an evaluation, and producing both plot-level summaries and population estimates. Future vignettes will cover more advanced topics, such as:

*   **Ratio Estimation**: Estimating ratios (e.g., Volume per Acre).
*   **Change Estimation**: Analyzing Growth, Removals, and Mortality.
*   **Custom Stratification**: Using custom strata for estimation.
