---
title: "Growth, Removals, and Mortality"
knitr:
  opts_chunk:
    echo: true
    warning: false
    message: false
---

Growth, removals, and mortality (GRM) refers to a set of estimation procedures
used by the FIA to quantify forest change. GRM is somewhat peculiar in that more
constraints on potential outputs exist compared to the simpler case of status
estimates. This is because GRM relies, in part, on the prediction of
midpoint values (i.e., values of an attribute in between field measurements),
and these predictions are only made for a subset of variables. Following this
philosophy, `fiaplyr` uses a specific analysis specification, `GRMAnalysis`,
that faithfully implements this paradigm. A more generic, but experimental,
form of change estimation is in development called `PairAnalysis` (forthcoming).
Still, `GRMAnalysis` is a powerful tool for estimating forest change, and
produces compliant estimates with a wide range of utility. This vignette
demonstrates the underlying GRM concepts, and shows how two examples:

1. Annual net growth of merchantable bole wood volume

## Specifying a Handler

To begin with GRM analyses, specify a handler as always. However, this handler
must use an evaluation ending in `3` which indicates a GRM-specific evaluation.
We also pass the `GRMAnalysis` specification.

```{r}
library(fiaplyr)
library(DBI)
library(duckdb)
library(dplyr)

con <- dbConnect(duckdb(), fiadb_vt_mini_path())
handler <- eval_handler(con, 501103, spec = new("GRMAnalysis"))

handler
```

Here, we see the handler is initialized with a GRM evaluation, and the summary
statement gives particulars. In this type of evaluation, the `tree` and `cond`
helpers refer to the revisit data. Additional slots that are not given in
a `StatusAnalysis` are available, which are used to derive change values,
among other uses. The most important of these tables is `tree_history`, which
represents the history of tree measurements for the beginning, midpoint, and
end of the remeasurement period. Users interact with this table using the
`tree_history()` helper, among other functions.

## GRM Transformations

Roughly, GRM analyses involve sophisticated transformations of columns in the
`tree_history` slot, and these transformations depend upon the objectives of the
user. To create these transformations, users rely on the `transform` function,
used in tandem with the `tree_history()` helper. For example, if the user
desires an estimate of the mortality volume over the period, then each tree is
assigned a value of zero if it is alive at the end of the period, and the
midpoint volume if it is dead at the end of the period. This can be done "by
hand" like so

```{r}
handler |>
  transform(
    tree_history(
      mort_vol = case_when(
        STATUSCD_begin == 1 & STATUSCD == 2 ~ VOL_begmid,
      )
    )
  ) 
```






A basic example is that of mortality. An initial volume
(or other attribute) is observed at the beginning of the remeasurement period,
and a final volume is observed at the end. If the tree died during the period,
then the contribution of the tree to

## Annual Net Growth of Merchantable Bole Wood Volume

Following the GRM paradigm, net growth represents the amount of added woody
material across the landscape after subtracting mortality (note that removals
are not subtracted). Dividing this value by the number of years in the
remeasurement period gives the annual net growth.

```{r}
handler |>

```

