---
title: "Basic Growth, Removals, and Mortality Analysis"
knitr:
  opts_chunk:
    echo: true
    warning: false
    message: false
---

Growth, removals, and mortality (GRM) refers to a set of estimation procedures
used by the FIA to quantify forest change. Broadly, trees are assigned to change
classes, which we refer to as transitions. Then, estimates are constructed based
on the transition and desired output of the user. Typical outputs include
estimates of mortality volume, removals, and others. Transitions are dependent
on a pre-specified `tree_basis` and `land_basis`. For example, ingrowth with
respect to growing stock is more stringent than ingrowth with respect to all
live trees, because growing stock requires a larger minimum diameter and more
strict defect constraints. Thus, users must specify the tree and land bases,
which will facilitate `fiaplyr` in correctly applying the appropriate transition
rules. Of course, users can entertain custom transitions as well, and such
advanced use cases are covered in [Advanced Growth, Removals, and Mortality]
(growth_removals_mortality_advanced.qmd).

## Specifying a Handler

To begin with GRM analyses, specify a handler as always. However, this handler
must use an evaluation ending in `3` which indicates a GRM-specific evaluation.
We also pass the `GRMAnalysis` specification.

```{r}
library(fiaplyr)
library(DBI)
library(duckdb)
library(dplyr)

con <- dbConnect(duckdb(), fiadb_vt_mini_path())
handler <- eval_handler(con, 501103, spec = grm_analysis())

handler
```

Here, we see the handler is initialized with a GRM evaluation, and the summary
statement gives particulars. The specification defaults to a tree basis of
`all_live` and a land basis of `forest_land`, but these can be changed by
setting the `tree_basis` and `land_basis` arguments in the `grm_analysis()`
function. Options for the tree basis include `all_live`, `growing_stock`, and
`sawtimber` and options for the land basis include `forest_land` and
`timberland`.

The core table involved with GRM analyses is the `tree_history` table, which
contains the tree-level histories for each tree in the remeasurement period, and
relies on the GRM paradigm of beginning, midpoint, and end values. Users can
view the lazy query for the `tree_history` table with

```{r}
handler@tables$tree_history |>
  select(DIA_begin, DIA_midpt, DIA) |>
  head()
```

Notice that the beginning and midpoint values have suffixes, while the end does
not. This complies with typical FIA conventions.

## Tree Transitions

In the GRM paradigm, trees undergo transitions, primarily based on their life
history. For example, a tree that is measured and alive at the beginning of the
period may survive to the endpoint, wherein it is classified as a survivor tree.
Tree transitions are applied when the handler is initialized, using transition
rules that comply with the selected tree and land bases. Users should interpret
transitions as a column in the `tree_history` table that contains the transition
type, for which we use the conventional names from the FIA database, and are
stored in the `transition` column.

```{r}
handler@tables$tree_history |>
  select(transition) |>
  head()
```

## Aggregation and Estimation of Change Components

Change components refer to the various outputs that users may be interested in,
such as mortality volume, removals, and others. For high level uses, users
obtain change components using scoped helpers within the `tree_history`
call, all of which have the `grm_` prefix. For example, to obtain net cubic
foot mortality volume, use the following

```{r}
ps <- PostStratifiedEstimator(handler |> partition(tree_history(transition)))

handler |>
  aggregate(
    tree_history(
      mort = grm_mortality(VOLCFSND)
    )
  )

ps |>
  estimate(
    tree_history(
      mort = grm_mortality(VOLCFSND, annualize = TRUE)
    ), output = "total"
  )
```

Other change components helpers include `grm_removals`, `grm_ingrowth`, and
`grm_net_change`, `grm_net_growth`, `grm_accretion` and others.
