---
title: "Growth, Removals, and Mortality"
knitr:
  opts_chunk:
    echo: true
    warning: false
    message: false
---

Growth, removals, and mortality (GRM) refers to a set of estimation procedures
used by the FIA to quantify forest change. Broadly, the paradigm relies on
three discrete time points, referred to as the beginning, midpoint, and end of
the remeasurement period for a given plot. Beginning and end values are
observed, or at least derived from observed values, and the midpoint values 
are predicted based on these observations. We refer to this three point
structure as a tree history, and it is the core of the GRM paradigm. In this
vignette, we demonstrate how to produce GRM estimates using `fiaplyr`, and
demonstrate that interacting with the tree history, which is a slot  within
the handler, allows for a wide range of change estimates.

## Specifying a Handler

To begin with GRM analyses, specify a handler as always. However, this handler
must use an evaluation ending in `3` which indicates a GRM-specific evaluation.
We also pass the `GRMAnalysis` specification.

```{r}
library(fiaplyr)
library(DBI)
library(duckdb)
library(dplyr)

con <- dbConnect(duckdb(), fiadb_vt_mini_path())
handler <- eval_handler(con, 501103, spec = new("GRMAnalysis"))

handler
```

Here, we see the handler is initialized with a GRM evaluation, and the summary
statement gives particulars. In this type of evaluation, the `tree` and `cond`
helpers refer to the revisit data. The `tree_history` slot represents the
history of tree measurements for the beginning, midpoint, and end of the
remeasurement period. Users interact with this table using the `tree_history()`
helper, much in the same way as regular status variables. The `tree_history`
table is expansive, containing hundreds of columns, but some basics can be
viewed to learn about the structure of the data. For example, to see the
beginning, midpoint, and end values for diameter, 

```{r}
handler@tables$tree_history |>
  select(INVYR, PLOT, DIA_begin, DIA_midpt, DIA) |>
  head()
```

Notice that the beginning and midpoint values have suffixes, while the end does
not. This complies with typical FIA conventions.

## Tree Transitions

In the GRM paradigm, trees undergo transitions, primarily based on their life
history. For example, a tree that is measured and alive at the beginning of the
period may survive to the endpoint, wherein it is classified as a survivor tree.
`fiaplyr` includes a suite of transition functions that ease the computation
of transition values, all of which have the `grm_` prefix. All transition
functions can be used in the `tree_history()` helper.

```{r}
handler@tables$tree_history |>
  mutate(
    survivor = grm_survivor(),
    mortality = grm_mortality(),
    ingrowth_live = grm_ingrowth_live()
  ) |>
  select(survivor, mortality, ingrowth_live) |>
  head()
```

## GRM Transformations

Roughly, GRM analyses involve sophisticated transformations of columns in the
`tree_history` slot, and these transformations depend upon the objectives of the
user. To create these transformations, users rely on the `transform` function,
used in tandem with the `tree_history()` helper. For example, if the user
desires an estimate of the mortality volume over the period, then each tree is
assigned a value of zero if it is alive at the end of the period, and the
midpoint volume if it is dead at the end of the period. This can be done using
the `grm_mortality()` helper

```{r}
handler |>
  transform(
    tree_history(
      survivor = grm_survivor()
    )
  ) 
```
    This estimate path is available for `GRMAnalysis` handlers that include the
    `tree_history` table.


## Aggregation the the Plot Level

When dealing with multiple plot visits, aggregation to the plot level is more
complex than the single date case. This occurs because the core operation of
making an aggregate is a weighted sum involving the expansion factor
`TPA_UNADJ`, and one must engage with the question of which expansion factor
to use: either the beginning, middle, or ending trees per acre. 


```{r}
mort_handler <- handler |>
  transform(
    tree_history(
      mortality = (STATUSCD_begin == 1L & STATUSCD == 2L) | (is.na(STATUSCD_begin) & STATUSCD == 2L),
      ingrowth = grm_ingrowth_live()
    )
  ) |>
  subset(
    tree_history(mortality == 1, DIA_begin >= 5, WOODLAND == 'N')
  ) |>
  transform(
    tree_history(volcfsnd_annual = VOLCFSND_midpt / REMPER)
  )

y_bar <- mort_handler |>
  aggregate(tree_history(volcfsnd_annual)) |>
  pull(volcfsnd_annual) |>
  mean()

A <- mort_handler@tables$pop_estn_unit |>
  pull(AREA_USED) |>
  sum()

A * y_bar

mort_handler |>
  PostStratifiedEstimator() |>
  estimate(tree_history(volcfsnd_annual), output = "total")
```



A basic example is that of mortality. An initial volume
(or other attribute) is observed at the beginning of the remeasurement period,
and a final volume is observed at the end. If the tree died during the period,
then the contribution of the tree to

## Annual Net Growth of Merchantable Bole Wood Volume

Following the GRM paradigm, net growth represents the amount of added woody
material across the landscape after subtracting mortality (note that removals
are not subtracted). Dividing this value by the number of years in the
remeasurement period gives the annual net growth.

```{r}
handler |>

```

