---
title: "Inventory Tables and Figures"
---

The FIA routinely produces tables and figures that display series of estimates
used for reporting purposes. `fiaplyr` is not itself a tabling or visualization
engine, rather we rely on pre-existing R packages to produce tables and figures.
This vignette serves as a cookbook for making tables with `gt` and figures with
`ggplot2`. Basic products can be made with relatively little effort. We produce
a table of estimates of growing stock by ownership and forest type, and a
figure of growing stock by county, as illustrations.

## A Crosstab of Growing Stock by Ownership and Forest Type

As always, load some prerequisite packages, establish a database connection, and
specify an `eval_handler`.

```{r}
#| label: setup
library(fiaplyr)
library(DBI)
library(duckdb)
library(dplyr)
library(tidyr)
library(gt)
library(ggplot2)

# Connect to the Vermont mini database
con <- dbConnect(duckdb(), fiadb_vt_mini_path())

# Create a handler for the 2003 to 2006 evaluation for Vermont status variables
handler <- eval_handler(con, 500601)
```

```{r}
#| label: estimate-growing-stock
gs_handler <- handler |>
  filter_tree(WOODLAND == 'N', TREECLCD == 2) |>
  set_cond_domains(OWNGRPCD, FORTYPCD)

ps <- PostStratifiedEstimator(gs_handler)

gs_ests <- estimate(ps, tree(VOLCFNET), margins = TRUE)
```