---
title: "Ratio Estimates"
knitr:
  opts_chunk:
    echo: true
    warning: false
    message: false
---

The oft overlooked ratio estimator is a powerful component of FIA analyses,
enabling the estimation of ratios that can normalize estimates to forested
areas, yield averages of tree-level quantities, estimate mortality rates,
and more. Users are encouraged to read the material of [@scott etc chapter] for
useful examples regarding the ratio estimator before proceeding. An
understanding of the more basic [status estimates](status_estimates.qmd)
vignette is also required.

As the name suggests, the ratio estimator is formed by dividing a numerator
estimate by a denominator estimate, making an analysis dependent on pairs of
variables. In `fiaplyr`, we accomplish this by passing two handler objects to
the estimator, one that specifies all estimates in the numerator, and another
that specifies all estimates in the denominator. Then, estimates for all
possible pairings are made. This can create an extremely expansive set of
estimates for the user to interact with, and can be overwhelming at times. To
reduce the complexity of the output, we suggest using filtering functions
(`filter_tree`, `filter_cond`, etc.), rather than relying on granular domains.

To illustrate the power of the ratio estimator, we will estimate three ratios
for the state of Vermont between 2003 and 2006.

## Growing Stock divided by Forested Area

Normalizing estimates by forested area is a common practice in FIA analyses.
For example, in states with sparse forest cover, per-acre densities that can
be made with `PostStratifiedEstimator` may seem small to the lay user because
the density is expressed over the entire state, rather than only for forests
lands within it. The ratio estimator resolves this by dividing the per-acre
density by the proportion of forested land, inflating the estimate to represent
the density exclusively within forests. This is straightforward using the
`PostStratifiedRatioEstimator`. First, establish a handler

```{r}
library(fiaplyr)
library(DBI)
library(duckdb)
library(dplyr)

# Connect to the Vermont mini database
con <- dbConnect(duckdb(), fiadb_vt_mini_path())

# Create a handler for the 2003 to 2006 evaluation for Vermont status variables
handler <- eval_handler(con, 500601)
```

Then, create two sub-handlers, one that constructs the numerator (growing stock)
and one that constructs the denominator (forested area).

```{r}
# Define the numerator handler for growing stock (TREECLCD == 2)
gs_handler <- handler |>
  filter_tree(TREECLCD == 2)

# Define the denominator handler for forested area (COND_STATUS_CD == 1)
fa_handler <- handler |>
  filter_cond(COND_STATUS_CD == 1)
```

Establish the `PostStratifiedRatioEstimator` using the two handlers, and then
use the `estimate_ratio` method with scoped helpers to define the numerator and
denominator targets of interest. In this case, we estimate `VOLCFNET` for the
tree numerator and use `cond()` for the denominator, which represents the
condition proportion.

```{r}
# Establish the psr estimator
psr <- PostStratifiedRatioEstimator(gs_handler, fa_handler)

gstk_ests <- estimate_ratio(psr, tree(VOLCFNET), cond())

gstk_ests
```

Here, we see the density of growing stock per acre on forested land, which is
1,807 cubic feet per acre. The table defines the numerator variable `var_n`,
and the denominator variable `var_d`. The point estimate and standard error are
provided.

## Diameter Height Ratio by Species

A representation of the slenderness of trees is the diameter-height ratio, which
has implications for taper, merchantability, and other characteristics of trees.
We will estimate the diameter height ratio for growing stock trees within the 10
to 15 inch diameter class for each species. By default, the ratio estimator
generates estimates for all possible domain pairings, but for this question
it is sufficient to only generate ratios when the species match between the
numerator and denominator, this can be done by using
`domain_pairing = 'matched'` argument.

```{r}
tree_handler <- handler |>
  filter_tree(TREECLCD == 2, DIA >= 10, DIA <= 15, ACTUALHT == HT) |>
  set_tree_domains(SPCD)

psr <- PostStratifiedRatioEstimator(tree_handler, tree_handler)

species_labels <- handler@tables$ref_species |>
  select(SPCD, COMMON_NAME) |>
  collect()

dhr <- estimate_ratio(psr, tree(HT), tree(DIA), include_components = TRUE, domain_pairing = "matched") |>
  arrange(desc(estimate)) |>
  filter(se != 0) |>
  left_join(species_labels, by = c("SPCD_n" = "SPCD"))

head(dhr)
```
