---
title: "Status Estimates"
knitr:
  opts_chunk:
    echo: true
    warning: false
    message: false
---

```{r include=FALSE}
db_path <- "../db/SQLite_FIADB_OR.db"
```

One of the most frequent needs of FIA data are estimates of status variables.
Status variables capture the state of a forest over a given window of time,
typically specified by the temporal range of an evaluation. In this vignette,
we will estimate various status variables for the state of Oregon using an
evaluation that covers the first cycle of the inventory, 2001 to 2010. We will
produce estimates for three use cases:

1. The proportion of forested land in the state.
2. Growing stock by species group.

## Establishing the Handler

As always, we need to specify a handler, which manages the connection to the
database for a specific evaluation. We will use the 2001 to 2010 evaluation for
Oregon status variables, which has evaluation code `411001`.

```{r}
library(fiaplyr)
library(DBI)
library(dplyr)

# Connect to the database
con <- dbConnect(RSQLite::SQLite(), db_path)

# Create a handler for the 2001 to 2010 evaluation for Oregon status variables
handler <- eval_handler(con, 411001)

# Print the handler summary
handler
```

## Proportion of Forested Land

The estimation of the proportion of forested land involves the use of
conditions, which are mapped portions of the plot. Further reading on conditions
can be found [here, todo]. If we are interested in the estimate of the
proportion of forested land, we can make a new handler that is filtered to only
include conditions that are forested, which is obtained by filtering on
`COND_STATUS_CD == 1`.

```{r}
forested_handler <- handler |>
  filter_cond(COND_STATUS_CD == 1)
```

Post-stratified estimates are made using the `PostStratifiedEstimator()` class.
This class implements the standard FIA post-stratified estimator used most
frequently in FIA state reports and similar products. Pass the handler to the
estimator, and specify `cond ~ 1` as the estimation formula, which is the
standard way to estimate condition proportions. The left-hand side of the
estimation formula specifies the slot to estimate, and the right-hand side
specifies the variables to use in the estimation. `1` is a placeholder used
for the basic case of proportions.

```{r}
ps_estimator <- PostStratifiedEstimator(forested_handler)

for_prop_est <- ps_estimator |>
  estimate(cond ~ 1)

for_prop_est
```

```{r include=FALSE}
est_vec <- for_prop_est |>
  pull(estimate)

sd_vec <- for_prop_est |>
  pull(se)
```

The proportion of forested land in Oregon between 2001 and 2010 was
`r est_vec[1]` $\pm$ `r sd_vec[1]`.

## Growing Stock by Species Group

The estimation of growing stock is frequently used to assess the amount of
merchantable timber within a region. Growing stock trees are identified in the
tree table with `TREECLCD == 2`.

```{r}
growing_stock_handler <- handler |>
  filter_tree(TREECLCD == 2) |>
  set_tree_domains(SPGRPCD)

ps_estimator <- PostStratifiedEstimator(growing_stock_handler)

growing_stock_est <- ps_estimator |>
  estimate(tree ~ VOLCFGRS | VOLCFNET)

growing_stock_est
```

```{r include=FALSE}
est_vec <- growing_stock_est |>
  pull(VOLCFGRS)

sd_vec <- growing_stock_est |>
  pull(VOLCFGRS_se)
```

The total volume of growing stock in Oregon between 2001 and 2010 was
`r est_vec[1]` $\pm$ `r sd_vec[1]` cubic feet per acre.