---
title: "Status Estimates"
knitr:
  opts_chunk:
    echo: true
    warning: false
    message: false
---

One of the most frequent needs of FIA data are estimates of status variables.
Status variables capture the state of a forest over a given window of time,
typically specified by the temporal range of an evaluation. In this vignette,
we will estimate various status variables for the state of Vermont using an
evaluation that covers the first cycle of the inventory, 2003 to 2006. This
inventory is included in a miniature database, and readers should be able to
follow along with their own installation

We will produce estimates for three use cases:

1. The proportion of forested land in the state.
2. Growing stock by species group.
3. Average diameter by species.

The first two cases are examples of post-stratified estimation, while the third
is involves the more complex ratio estimator. The ratio estimator is covered
more thouroughly in a separate vignette.

## Establishing the Handler

As always, we need to specify a handler, which manages the connection to the
database for a specific evaluation. We will use the 2003 to 2006 evaluation for
Vermont status variables, which has evaluation code `500601`.

```{r}
library(fiaplyr)
library(DBI)
library(duckdb)
library(dplyr)

# Connect to the Vermont mini database
con <- dbConnect(duckdb(), fiadb_vt_mini_path())

# Create a handler for the 2003 to 2006 evaluation for Vermont status variables
handler <- eval_handler(con, 500601)

# Print the handler summary
handler
```

## Proportion of Forested Land

The estimation of the proportion of forested land involves the use of
conditions, which are mapped portions of the plot. Further reading on conditions
can be found [here, todo]. If we are interested in the estimate of the
proportion of forested land, we can make a new handler that is filtered to only
include conditions that are forested, which is obtained by filtering on
`COND_STATUS_CD == 1`.

```{r}
forested_handler <- handler |>
  filter_cond(COND_STATUS_CD == 1)
```

Post-stratified estimates are made using the `PostStratifiedEstimator()` class.
This class implements the standard FIA post-stratified estimator used most
frequently in FIA state reports and similar products. Pass the handler to the
estimator, and specify `cond ~ 1` as the estimation formula, which is the
standard way to estimate condition proportions. The left-hand side of the
estimation formula specifies the slot to estimate, and the right-hand side
specifies the variables to use in the estimation. `1` is a placeholder used
for the basic case of proportions.

```{r}
ps_estimator <- PostStratifiedEstimator(forested_handler)

for_prop_est <- ps_estimator |>
  estimate(cond ~ 1)

for_prop_est
```

```{r include=FALSE}
est_vec <- for_prop_est |>
  pull(estimate)

sd_vec <- for_prop_est |>
  pull(se)
```

## Growing Stock by Species Group

The estimation of growing stock is frequently used to assess the amount of
merchantable timber within a region. Growing stock trees are identified in the
tree table with `TREECLCD == 2`.

```{r}
growing_stock_handler <- handler |>
  filter_tree(TREECLCD == 2) |>
  set_tree_domains(SPGRPCD)

ps_estimator <- PostStratifiedEstimator(growing_stock_handler)

growing_stock_est <- ps_estimator |>
  estimate(tree ~ VOLCFGRS | VOLCFNET)

View(growing_stock_est)
```